---
title: "DPGLM Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cdpglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The `dirglm` package provides various Bayesian semi-parametric extensions to the generalized linear model. These are
available for use through the functions `dirglm`, `dpglm`, and `cdpglm`.

In this document, we will explore how to use the `cdpglm` function and inspect its results. In particular, we will

1. Simulate longitudinal data from an exponential-tilting model with a Gaussian copula dependence structure
2. Fit `cdpglm`
3. Summarize and visualize the results

```{r setup}
library(dirglm)
```

# Simulate Data

```{r}
set.seed(832)

n_subjects <- 20
n_visits <- 4
n <- n_subjects * n_visits

subject <- rep(seq_len(n_subjects), each = n_visits)
visit <- rep(seq(-0.5, 0.5, length.out = n_visits), times = n_subjects)
x <- rnorm(n_subjects)
x <- rep(x, each = n_visits)

X <- model.matrix(~ x + visit)
beta_true <- c(0.2, 0.5, -0.4)

mu <- plogis(drop(X %*% beta_true))

# Continuous-like baseline on (0, 1)
spt <- seq(1e-4, 1 - 1e-4, length.out = 201)
f0 <- dnorm(spt, mean = 0.55, sd = 0.15)
f0 <- f0 / sum(f0)

theta <- gldrm:::getTheta(
  spt = spt,
  f0 = f0,
  mu = mu,
  sampprobs = NULL,
  ySptIndex = NULL,
  thetaStart = NULL
)$theta

prob_mat <- t(vapply(seq_len(n), function(i) {
  w <- exp(theta[i] * spt) * f0
  w / sum(w)
}, numeric(length(spt))))

rho_true <- 0.5

Sigma <- matrix(rho_true, nrow = n_visits, ncol = n_visits)
diag(Sigma) <- 1

y <- numeric(n)

for (id in seq_len(n_subjects)) {
  idx <- which(subject == id)

  u <- stats::pnorm(drop(mvtnorm::rmvnorm(1, sigma = Sigma)))

  for (j in seq_along(idx)) {
    obs <- idx[j]
    cdf_vals <- cumsum(prob_mat[obs, ])
    y[obs] <- spt[which(cdf_vals >= u[j])[1]]
  }
}

dat <- data.frame(
  y = y,
  x = x,
  visit = visit,
  subject = subject
)
```

# Fit CDPGLM

Now, we will fit the `cdpglm` model. The number of MCMC draws is kept small to speed up computation, but practitioners may require more iterations for analysis.

```{r}
fit <- cdpglm(
  y ~ x + visit,
  data = dat,
  group_index = dat$subject,
  link = "logit",
  cdpglmControl = cdpglm.control(
    burnin = 10,
    thin = 1,
    save = 50,
    spt = c(0, 1),
    corr = "ex",
    seed = 832
  )
)
```

# Posterior Summary

The `summary()` method reports posterior summaries for the regression coefficients `beta`, reference distribution `crm`, and correlation parameter `rho`.

```{r}
summary(fit)
```

## Plotting

After fitting `cdpglm`, one may use `plot_cdpglm` to visually inspect some posterior summaries. We will show two examples. The first one inspects `beta`.

```{r}
plot_cdpglm(fit)
```

Similarly, we can inspect `rho`:

```{r}
plot_cdpglm(fit, what = "rho")
```

Finally, we can inspect the reference distribution. As with `plot_dirglm` and `plot_cdpglm`, users can customize some aesthetic aspects of the plot. We show another example here.

```{r}
plot_cdpglm(fit, what = "crm", col = "#56B4E9",
            main = "CRM Posterior Mean (95% Credible Interval)",
            xlab = "Atoms", ylab = "density")
```
