---
title: "dirglm_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dirglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dirglm)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will provide a few simulation examples for the `dirglm` function. Each function named `sim_dat_X` generates a data set corresponding to different instances of $\beta$, $f_0$, the link function, or some combination of those three. In what follows, each code section generates a data set from 

# Data-Generating Functions
```{r}
# Required Packages ----

require(mvtnorm)
require(gldrm)
require(extraDistr)
require(dplyr)

# Simulation scenario 1 ----
sim_dat_1 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  f0 <- extraDistr::dtpois(spt, lambda = 1, b = 5)     # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                       # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}


# Simulation scenario 2 ----
sim_dat_2 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  temp <- dpois(spt, lambda = 1)
  f0 <- numeric(6)
  f0[1] <- 3 * temp[1] / (3 * temp[1] + sum(temp[-1]))
  f0[-1] <- temp[-1] / (3 * temp[1] + sum(temp[-1]))  # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                        # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}

# Simulation scenario 3 ----
##  Multinomial logistic regression
sim_dat_3 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-1, 1)               # True beta
  spt <- 0:5
  mu <- 5 * plogis(X %*% beta)             # True mu
  f0  <- dbinom(spt, size = 5, prob = 0.5) # True f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}

# Simulation scenario 4 ----
##  Uniform baseline with identity link
sim_dat_4 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), runif(n * (p-1), min=-1, max=1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(5, 2)               # True beta
  mu <- X %*% beta                   # True mu
  spt <- 0:10
  l <- length(spt)
  f0 <- rep(1/l, l)                  # True f0
  theta <- gldrm:::getTheta(
    spt       = spt,
    f0        = f0,
    mu        = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}
```

# Simulation

## Example 1: Truncated Poisson
```{r}
## Example 1
spt <- 0:5
mu0 <- 1
data <- sim_dat_1(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, dirglmControl=dirglm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is truncated Poisson = (0.368, 0.368, 0.184, 0.061, 0.015, 0.003) [rounded]
```

## Example 2: Zero-inflated truncated Poisson
```{r}
## Example 2
spt <- 0:5
mu0 <- 1
data <- sim_dat_2(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, dirglmControl=dirglm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is tilted, truncated Poisson = (0.636, 0.212, 0.106, 0.035, 0.009, 0.002) [rounded]
```

## Example 3: Multiple Logistic Regression
```{r}
## Example 3
spt <- 0:5
mu0 <- 2.5
data <- sim_dat_3(n = 500, p = 3)
### Define link
link <- list(
  linkfun <- function(mu) log(mu / (5 - mu)),
  linkinv <- function(eta) 5 * plogis(eta),
  mu.eta  <- function(eta) 5 * plogis(eta) * (1 - plogis(eta))
)
names(link) <- c("linkfun", "linkinv", "mu.eta")
dir_glm.fit <- dirglm(y ~ X2 + X3, data=data, link=link,
                      dirglmControl=dirglm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-1, 1)
colMeans(f0_samples)          # true f0 is binomial = (0.031, 0.156, 0.313, 0.313, 0.156, 0.031) [rounded]
```

## Example 4: Uniform baseline with Identity link
```{r}
## Example 4
spt <- 0:10
mu0 <- 5
data <- sim_dat_4(n = 500, p = 5)
dir_glm.fit <- dirglm(y ~ X2 + X3 + X4 + X5, data=data, link="identity",
                        dirglmControl=dirglm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(5, 2)
colMeans(f0_samples)          # true f0 is uniform = (0.1, 0.1, 0.1, ..., 0.1)
```

### Example 4 ctd: Calculating frequentist coverage
```{r}
spt <- 0:10
l <- length(spt)
mu0 <- 5
beta_true <- c(5, 2, 0, 0, 0)  # True beta
f0_true <- rep(1/l, l) # True f0
p <- 5

beta_U <- beta_true + 0.25 * pmax(abs(beta_true), 1) # Since some beta = 0
beta_L <- beta_true - 0.25 * pmax(abs(beta_true), 1) # Since some beta = 0
f0_U <- f0_true + 0.25 * 1 / l
f0_L <- f0_true - 0.25 * 1 / l
M <- 500
mcmc <- 1000
p_ml_b <- matrix(NA, nrow=M, ncol=p)
p_mu_b <- matrix(NA, nrow=M, ncol=p)
p_ml_f0 <- matrix(NA, nrow=M, ncol=l)
p_mu_f0 <- matrix(NA, nrow=M, ncol=l)
all_samples_b <- matrix(NA, nrow=M*mcmc, ncol=p)
all_samples_f0 <- matrix(NA, nrow=M*mcmc, ncol=l)
for (m in 1:M) {
  if (m == 1 || m %% 50 == 0) cat(sprintf("Starting dirGLM fit #%d\n", m))
  # Generate data
  data <- sim_dat_4(n = 500, p = 5)
  dir_glm.fit <- dirglm(y ~ X2 + X3 + X4 + X5, data=data, link="identity",
                        dirglmControl=dirglm.control(mu0=mu0, spt=spt, save=mcmc))
  beta_samples <- dir_glm.fit$samples$beta
  f0_samples <- dir_glm.fit$samples$f0
  
  # Save samples
  all_samples_b[((m-1)*mcmc + 1):(m*mcmc), ] <- beta_samples
  all_samples_f0[((m-1)*mcmc + 1):(m*mcmc), ] <- f0_samples
  
  # Go over each beta
  for (bi in 1:p) {
    p_ml_b[m, bi] <- mean(beta_samples[, bi] >= beta_L[bi])
    p_mu_b[m, bi] <- mean(beta_samples[, bi] <= beta_U[bi])
  }
  
  # Go over each f0
  for (f0i in 1:l) {
    p_ml_f0[m, f0i] <- mean(f0_samples[, f0i] >= f0_L[f0i])
    p_mu_f0[m, f0i] <- mean(f0_samples[, f0i] <= f0_U[f0i])
  }
}
cat(sprintf("Done with dirGLM fits\n"))

# Compute average of pml and pmu
p_ml_bar_beta <- colMeans(p_ml_b)
p_mu_bar_beta <- colMeans(p_mu_b)
p_ml_bar_f0 <- colMeans(p_ml_f0)
p_mu_bar_f0 <- colMeans(p_mu_f0)
cat(sprintf("Computed p_bars\n"))

# Compute qml and qmu
m <- 1
q_ml_b <- matrix(NA, nrow=M, ncol=p)
q_mu_b <- matrix(NA, nrow=M, ncol=p)
q_ml_f0 <- matrix(NA, nrow=M, ncol=l)
q_mu_f0 <- matrix(NA, nrow=M, ncol=l)
for (m in 1:M) {
  idx_start <- (m - 1) * mcmc + 1
  idx_end <- m * mcmc
  cur_samples_b <- all_samples_b[idx_start:idx_end, ]   # mcmc x p
  cur_samples_f0 <- all_samples_f0[idx_start:idx_end, ] # mcmc x l

  for (bi in 1:p) {
    q_ml_b[m, bi] <- quantile(cur_samples_b[, bi], probs = p_ml_bar_beta[bi])
    q_mu_b[m, bi] <- quantile(cur_samples_b[, bi], probs = p_mu_bar_beta[bi])
  }
  for (f0i in 1:l) {
    q_ml_f0[m, f0i] <- quantile(cur_samples_f0[, f0i], probs = p_ml_bar_f0[f0i])
    q_mu_f0[m, f0i] <- quantile(cur_samples_f0[, f0i], probs = p_mu_bar_f0[f0i])
  }
}
cat(sprintf("Computed q_ms\n"))

# Compute "coverage"
s_l_beta <- rep(NA, p)
s_u_beta <- rep(NA, p)
s_l_f0 <- rep(NA, l)
s_u_f0 <- rep(NA, l)

# Go over each beta
for (bi in 1:p) {
  s_l_beta[bi] <- mean(q_ml_b[, bi] >= beta_L[bi])
  s_u_beta[bi] <- mean(q_mu_b[, bi] <= beta_U[bi])
}

# Go over each f0
for (f0i in 1:l) {
  s_l_f0[f0i] <- mean(q_ml_f0[, f0i] >= f0_L[f0i])
  s_u_f0[f0i] <- mean(q_mu_f0[, f0i] <= f0_U[f0i])
}

# Print results
cat("s_l_beta:", s_l_beta, "| p_l-bar:", p_ml_bar_beta, "\n")
cat("s_u_beta:", s_u_beta, "| p_u-bar:", p_mu_bar_beta, "\n")
cat("s_l_f0:", s_l_f0, "| p_l-bar:", p_ml_bar_f0, "\n")
cat("s_u_f0:", s_u_f0, "| p_u-bar:", p_mu_bar_f0, "\n")
```
