---
title: "bspgldrm_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bspgldrm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will provide a few simulation examples for the`bspgldrm` function. Each function named `sim_dat_X` generates a data set corresponding to different instances of $\beta$, $f_0$, the link function, or some combination of those three. In what follows, each code section generates a data set from 

# Data-Generating Functions
```{r}
# Required Packages ----

require(mvtnorm)
require(gldrm)
require(extraDistr)
require(dplyr)

# Simulation scenario 1 ----
sim_dat_1 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  f0 <- extraDistr::dtpois(spt, lambda = 1, b = 5)     # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                       # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}


# Simulation scenario 2 ----
sim_dat_2 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  temp <- dpois(spt, lambda = 1)
  f0 <- numeric(6)
  f0[1] <- 3 * temp[1] / (3 * temp[1] + sum(temp[-1]))
  f0[-1] <- temp[-1] / (3 * temp[1] + sum(temp[-1]))  # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                        # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}

# Simulation scenario 3 ----
##  Multinomial logistic regression
sim_dat_3 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-1, 1)               # True beta
  spt <- 0:5
  mu <- 5 * plogis(X %*% beta)             # True mu
  f0  <- dbinom(spt, size = 5, prob = 0.5) # True f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}

sim_dat_4 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), runif(n * (p-1), min=-1, max=1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(5, 2)               # True beta
  mu <- X %*% beta                   # True mu
  spt <- 0:10
  l <- length(spt)
  f0 <- rep(1/l, l)                  # True f0
  theta <- gldrm:::getTheta(
    spt       = spt,
    f0        = f0,
    mu        = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}
```

# Simulation

## Example 1: Truncated Poisson
```{r}
## Example 1
spt <- 0:5
mu0 <- 1
data <- sim_dat_1(n = 500, p = 2)
dir_glm.fit <- bspgldrm(y ~ X2, data=data, bspgldrmControl=bspgldrm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is truncated Poisson = (0.368, 0.368, 0.184, 0.061, 0.015, 0.003) [rounded]
```

## Example 2: Zero-inflated truncated Poisson
```{r}
## Example 2
spt <- 0:5
mu0 <- 1
data <- sim_dat_2(n = 500, p = 2)
dir_glm.fit <- bspgldrm(y ~ X2, data=data, bspgldrmControl=bspgldrm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is tilted, truncated Poisson = (0.636, 0.212, 0.106, 0.035, 0.009, 0.002) [rounded]
```

## Example 3: Multiple Logistic Regression
```{r}
## Example 3
spt <- 0:5
mu0 <- 2.5
data <- sim_dat_3(n = 500, p = 3)
### Define link
link <- list(
  linkfun <- function(mu) log(mu / (5 - mu)),
  linkinv <- function(eta) 5 * plogis(eta),
  mu.eta  <- function(eta) 5 * plogis(eta) * (1 - plogis(eta))
)
names(link) <- c("linkfun", "linkinv", "mu.eta")
dir_glm.fit <- bspgldrm(y ~ X2 + X3, data=data, link=link,
                        bspgldrmControl=bspgldrm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-1, 1)
colMeans(f0_samples)          # true f0 is binomial = (0.031, 0.156, 0.313, 0.313, 0.156, 0.031) [rounded]
```

## Example 4: Uniform baseline with Identity link
```{r}
## Example 4
spt <- 0:10
mu0 <- 5
data <- sim_dat_4(n = 500, p = 5)
dir_glm.fit <- bspgldrm(y ~ X2 + X3 + X4 + X5, data=data, link="identity",
                        bspgldrmControl=bspgldrm.control(mu0=mu0, spt=spt))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(5, 2)
colMeans(f0_samples)          # true f0 is uniform = (0.1, 0.1, 0.1, ..., 0.1)
```


```{r}
simData <- function(n, p, betaMax, link, ySim) {
  beta <- c(0, seq(-betaMax, betaMax, length.out = p))
  X    <- cbind(1, matrix(rnorm(n * p), nrow = n))
  colnames(X) <- paste0("X", seq_len(ncol(X)))
  eta  <- as.vector(X %*% beta)
  mu   <- link$linkinv(eta)
  y    <- ySim(n, mu)
  list(x = X, y = y, eta = eta, mu = mu, beta = beta)
}

set.seed(100)
ySim <- function(n, mu) rpois(n, 1)
lf <- stats::make.link("identity")
sim <- simData(n=100, p=0, betaMax=0, link=lf, ySim=ySim)
data <- data.frame(sim$x, y=sim$y)

m1 <- as.vector(table(data$y)) / length(data$y)
m2 <- bspgldrm(y ~ X1 - 1, data=data, link="identity",
               bspgldrmControl=bspgldrm:::bspgldrm.control(burnin=1000,
                                                           thin=20,
                                                           save=10000))  # link function doesn't matter with no covariates

# Compare beta to mean(y)
tol        <- 0.1 * abs(beta_truth)
gamma      <- 0.95
iter       <- m2$iter
beta_truth <- mean(data$y)
beta_model <- m2$samples$beta
L          <- beta_truth - tol
U          <- beta_truth + tol
cov_beta   <- sum( (L <= beta_model) & (beta_model <= U) ) / iter
cat(paste0("beta ", cov_beta, "\n"))

tol     <- 0.1 * abs(f0_truth)
f0_truth <- m1
f0_model <- m2$samples$f0
L        <- f0_truth - tol
U        <- f0_truth + tol
for (i in seq_along(f0_truth)) {
  cov_f0i <- sum( (L[i] <= f0_model[, i]) & (f0_model[, i] <= U[i]) ) / iter
  cat(paste0("f0 ", cov_f0i))
}
```
