---
title: "Dir-GLM Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dirglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The `dirglm` package provides various Bayesian semi-parametric extensions to the generalized linear model. These are
available for use through the functions `dirglm`, `dpglm`, and `cdpglm`.

In this document, we will explore how to use the `dirglm` function and inspect its results. In particular, we will

1. Simulate a simple regression model with discrete response on finite support
2. Fit `dirglm`
3. Summarize and visualize the results

```{r setup}
library(dirglm)
```

# Simulate Data

We will begin by simulating data from a simple binomial response model.

```{r}
set.seed(832)

n <- 100
m <- 5

x <- rnorm(n)
beta <- c(-0.25, 0.75) # True beta
eta <- beta[1] + beta[2] * x
p <- plogis(eta)
y <- rbinom(n, size = m, prob = p) / m

dat <- data.frame(y = y, x = x)
```

# Fit Dir-GLM

Now, we will fit the `dirglm` model. The number of MCMC draws is kept small to speed up computation, but practitioners may require more iterations for analysis.

```{r}
fit <- dirglm(
  y ~ x,
  data = dat,
  link = "logit",
  dirglmControl = dirglm.control(
    burnin = 10,
    thin = 1,
    save = 500,
    seed = 832
  )
)
```

# Posterior Summary

The `summary()` method reports posterior summaries for the regression coefficients `beta` and reference distribution `f0`.

```{r}
summary(fit)
```

## Plotting

After fitting `dirglm`, one may use `plot_dirglm` to visually inspect some posterior summaries. We will show two examples. The first one inspects `beta`.

```{r}
plot_dirglm(fit, what = "beta")
```

One can modify this plot in a few common ways. In the below chunk, we inspect the same posterior samples for `beta`, but make the following changes for visualization:
1. By setting `pars = "Intercept`, only the posterior draws for the intercept term are shown.
2. `prob = 0.9` displays a $90\%$ credible interval, as opposed to the default $95\%$.
3. `robust = TRUE` forces the plot to use the posterior median for computation instead of the posterior mean.
4. `col = "blue"` changes the default coloring used in the plot.

```{r}
plot_dirglm(fit, what = "beta", pars = "Intercept", prob = 0.9, robust = TRUE, col = "#E69F00")
```

Similarly, one can visualize reference distribution $f_0$. Similar modifications can be made to the plot. 

```{r}
plot_dirglm(fit, what = "f0", prob = 0.8, robust = TRUE, col = "#56B4E9")
```
