---
title: "dirglm_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dirglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dirglm)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will provide a few simulation examples for the `dirglm` function. Each function named `sim_dat_X` generates a data set corresponding to different instances of $\beta$, $f_0$, the link function, or some combination of those three. In what follows, each chunk generates a data set from a `sim_dat_X`, fits `dirglm`, and shows the relevant output.

# Data-Generating Functions
```{r}
# Required Packages ----

require(mvtnorm)
require(gldrm)
require(extraDistr)
require(dplyr)

# Simulation scenario 1 ----
sim_dat_1 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  f0 <- extraDistr::dtpois(spt, lambda = 1, b = 5)     # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                       # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}


# Simulation scenario 2 ----
sim_dat_2 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  temp <- dpois(spt, lambda = 1)
  f0 <- numeric(6)
  f0[1] <- 3 * temp[1] / (3 * temp[1] + sum(temp[-1]))
  f0[-1] <- temp[-1] / (3 * temp[1] + sum(temp[-1]))  # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                        # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}

# Simulation scenario 3 ----
##  Multinomial logistic regression
sim_dat_3 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-1, 1)               # True beta
  spt <- 0:5
  mu <- 5 * plogis(X %*% beta)             # True mu
  f0  <- dbinom(spt, size = 5, prob = 0.5) # True f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}

# Simulation scenario 4 ----
##  Uniform baseline with identity link
sim_dat_4 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), runif(n * (p-1), min=-1, max=1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(5, 2)               # True beta
  mu <- X %*% beta                   # True mu
  spt <- 0:10
  l <- length(spt)
  f0 <- rep(1/l, l)                  # True f0
  theta <- gldrm:::getTheta(
    spt       = spt,
    f0        = f0,
    mu        = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}
```

# Simulation

## Example 1: Truncated Poisson
```{r}
## Example 1
spt <- 0:5
mu0 <- 1
data <- sim_dat_1(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is truncated Poisson = (0.368, 0.368, 0.184, 0.061, 0.015, 0.003) [rounded]
```

## Example 2: Zero-inflated truncated Poisson
```{r}
## Example 2
spt <- 0:5
mu0 <- 1
data <- sim_dat_2(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, type="dirglm", dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is tilted, truncated Poisson = (0.636, 0.212, 0.106, 0.035, 0.009, 0.002) [rounded]
```

## Example 3: Multiple Logistic Regression
```{r}
## Example 3
spt <- 0:5
mu0 <- 2.5
data <- sim_dat_3(n = 500, p = 3)
### Define link
link <- list(
  linkfun <- function(mu) log(mu / (5 - mu)),
  linkinv <- function(eta) 5 * plogis(eta),
  mu.eta  <- function(eta) 5 * plogis(eta) * (1 - plogis(eta))
)
names(link) <- c("linkfun", "linkinv", "mu.eta")
dir_glm.fit <- dirglm(y ~ X2 + X3, data=data, link=link, type="dirglm",
                      dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-1, 1, 0)
colMeans(f0_samples)          # true f0 is binomial = (0.031, 0.156, 0.313, 0.313, 0.156, 0.031) [rounded]
```

## Example 4: Uniform baseline with Identity link
```{r}
## Example 4
spt <- 0:10
mu0 <- 5
data <- sim_dat_4(n = 500, p = 5)
dir_glm.fit <- dirglm(y ~ X2 + X3 + X4 + X5, data=data, link="identity", type="dirglm",
                        dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(5, 2, 0, 0, 0)
colMeans(f0_samples)          # true f0 is uniform = (0.1, 0.1, 0.1, ..., 0.1)
```

# DPGLM
```{r}
rm(list = ls())
devtools::load_all()

## --------------------------
## Simulation III setup
## --------------------------
f0_true_kde <- readRDS("../DPGLM/DPGLM/sim_truth/f0_kde_unif_0_1.rds")
spt   <- f0_true_kde$spt
f0_kde <- f0_true_kde$f0

dx <- diff(spt)[1]
m0 <- sum(spt * f0_kde) * dx

truth1 <- list(
  beta = c(1, 0),
  f0   = matrix(c(spt, f0_kde), ncol = 2, byrow = FALSE)
)

# truth2 <- list(
#   beta = c(0.2, 0.7),
#   f0   = matrix(c(spt, f0_kde), ncol = 2, byrow = FALSE)
# )

## --------------------------
## Data generator
## --------------------------
sim_data <- function(n, true_beta) {
  X <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])

  mu <- as.numeric(stats::plogis(X %*% true_beta))

  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta

  # Compute the matrix of exponentiated outer products
  exp_matrix <- exp(outer(theta, spt, "*"))

  # Compute the dot product row-wise and apply the log
  btheta <- log(exp_matrix %*% f0_kde * dx)

  X <- as.matrix(X)
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })

  return(data.frame(Y, theta = theta, X))
}

# Run simulation
set.seed(832)
iter <- 20
dat <- sim_data(100, truth1$beta)
truth1$f0[, 2] <- truth1$f0[, 2] / sum(truth1$f0[, 2])
fit <- dpglm(Y ~ X2, data = data.frame(Y = dat$Y, X2 = dat$X2), link = "logit",
             dpglmControl = dpglm.control(burnin = 10, save = iter, thin = 1,
                                          seed = 832))
```

```{r}
crm_metrics <- function(zlist, Jlist) {
  stopifnot(length(zlist) == length(Jlist))
  T <- length(zlist)

  out <- data.frame(
    iter = seq_len(T),
    n_atoms = NA_integer_,
    z_min = NA_real_,
    z_max = NA_real_,
    J_sum = NA_real_,
    J_max = NA_real_,
    J_max_share = NA_real_,
    ess = NA_real_
  )

  for (t in seq_len(T)) {
    z <- zlist[[t]]
    J <- Jlist[[t]]
    if (is.null(z) || is.null(J) || length(z) == 0 || length(J) == 0) next

    Jpos <- pmax(J, 0)
    sJ <- sum(Jpos)
    w <- if (sJ > 0) Jpos / sJ else rep(NA_real_, length(Jpos))

    out$n_atoms[t]     <- length(z)
    out$z_min[t]       <- min(z)
    out$z_max[t]       <- max(z)
    out$J_sum[t]       <- sJ
    out$J_max[t]       <- max(Jpos)
    out$J_max_share[t] <- if (sJ > 0) max(w) else NA_real_
    out$ess[t]         <- if (sJ > 0) 1 / sum(w^2) else NA_real_
  }

  out
}

summ_num <- function(x, name = deparse(substitute(x))) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(data.frame(name=name, n=0, min=NA, q05=NA, med=NA, mean=NA, q95=NA, max=NA))
  data.frame(
    name = name,
    n    = length(x),
    min  = min(x),
    q05  = unname(stats::quantile(x, 0.05)),
    med  = stats::median(x),
    mean = mean(x),
    q95  = unname(stats::quantile(x, 0.95)),
    max  = max(x)
  )
}

# Beta
new_beta_mat <- fit$samples$beta
new_beta_mean <- colMeans(new_beta_mat, na.rm = TRUE)
new_beta_sd   <- apply(new_beta_mat, 2, sd, na.rm = TRUE)

print(data.frame(beta_mean = new_beta_mean, beta_sd = new_beta_sd))
cat("beta_acceptance =", fit$beta_acceptance, "\n")
cat("crm_acceptance  =", fit$crm_acceptance, "\n")

# CRM
new_zlist <- fit$samples$crm$z.tld
new_Jlist <- fit$samples$crm$J.tld
new_crm_df <- crm_metrics(new_zlist, new_Jlist)

cat("\n--- CRM summary---\n")
print(rbind(
  summ_num(new_crm_df$n_atoms, "n_atoms"),
  summ_num(new_crm_df$ess, "ess"),
  summ_num(new_crm_df$J_max_share, "J_max_share"),
  summ_num(new_crm_df$z_min, "z_min"),
  summ_num(new_crm_df$z_max, "z_max")
))

# End behavior
cat("\n--- Last 10 CRM iters ---\n")
print(tail(new_crm_df, 5))
```

# CDPGLM

```{r}
make_truth_grid <- function() {
  truth <- readRDS("../..//f0_kde_unif_0_1.rds")
  spt <- as.numeric(truth$spt)
  f0 <- as.numeric(truth$f0)
  dx <- diff(spt)[1]
  f0 <- f0 / sum(f0 * dx)

  list(
    spt = spt,
    f0 = f0,
    dx = dx
  )
}

make_longitudinal_design <- function(n_subjects, n_visits) {
  visit_values <- if (n_visits == 1L) {
    0
  } else {
    seq(-0.5, 0.5, length.out = n_visits)
  }

  subject <- rep(seq_len(n_subjects), each = n_visits)
  biomarker_subject <- 2L * rbinom(n_subjects, size = 1L, prob = 0.5) - 1L
  biomarker <- rep(biomarker_subject, each = n_visits)
  visit <- rep(visit_values, times = n_subjects)

  data.frame(
    subject = subject,
    biomarker = biomarker,
    visit = visit,
    row.names = NULL
  )
}

simulate_et_data <- function(n_subjects, n_visits, true_beta, rho, truth_grid, seed = 832) {
  set.seed(seed)

  design <- make_longitudinal_design(n_subjects = n_subjects, n_visits = n_visits)
  X <- model.matrix(~ biomarker * visit, data = design)
  mu <- plogis(drop(X %*% true_beta))

  theta <- gldrm:::getTheta(
    spt = truth_grid$spt,
    f0 = truth_grid$f0,
    mu = mu,
    thetaStart = NULL,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta

  exp_matrix <- exp(outer(theta, truth_grid$spt, "*"))
  btheta <- log(drop(exp_matrix %*% truth_grid$f0) * truth_grid$dx)

  mass_matrix <- t(vapply(seq_len(nrow(X)), function(i) {
    weights <- exp(theta[i] * truth_grid$spt - btheta[i]) * truth_grid$f0 * truth_grid$dx
    weights / sum(weights)
  }, numeric(length(truth_grid$spt))))

  # Exchangeable copula
  Sigma <- matrix(rho, nrow = n_visits, ncol = n_visits)
  diag(Sigma) <- 1

  y <- numeric(nrow(design))
  for (subject_id in seq_len(n_subjects)) {
    idx <- which(design$subject == subject_id)
    u <- pnorm(drop(mvtnorm::rmvnorm(1, sigma = Sigma)))

    for (j in seq_along(idx)) {
      obs <- idx[j]
      cdf_vals <- cumsum(mass_matrix[obs, ])
      hit <- which(cdf_vals >= u[j])[1]
      if (is.na(hit)) {
        hit <- length(truth_grid$spt)
      }
      y[obs] <- truth_grid$spt[hit]
    }
  }

  transform(
    design,
    y = pmin(pmax(y, 1e-6), 1 - 1e-6),
    mu_true = mu,
    theta_true = theta
  )
}
```

Fit cdpglm
```{r}
# Fit
devtools::load_all()
library(gldrm)
library(mvtnorm)

truth_grid <- make_truth_grid()

set.seed(832)
sim_dat <- simulate_et_data(
  n_subjects = 20,
  n_visits = 5,
  true_beta = c(0.6, -1.0, -0.8, 0.8),
  rho = 0.6,
  truth_grid = truth_grid,
  seed = 832
)

cdpglm.fit <- cdpglm(
  y ~ biomarker * visit,
  data = sim_dat,
  group_index = sim_dat$subject,
  link = "logit",
  cdpglmControl = cdpglm.control(
    save = 50,
    thin = 1,
    burnin = 10,
    spt = c(0, 1),
    seed = 832,
    corr = "ex"
  )
)
```

Checks
```{r}
## --------------------------
## CDPGLM diagnostics
## --------------------------

summ_num <- function(x, name) {
  data.frame(
    name = name,
    n = length(x),
    min = min(x, na.rm = TRUE),
    q05 = quantile(x, 0.05, na.rm = TRUE),
    med = median(x, na.rm = TRUE),
    mean = mean(x, na.rm = TRUE),
    q95 = quantile(x, 0.95, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    row.names = NULL
  )
}

crm_metrics <- function(zlist, Jlist) {
  out <- lapply(seq_along(zlist), function(i) {
    z <- zlist[[i]]
    J <- Jlist[[i]]
    Jsum <- sum(J)
    Jnorm <- J / Jsum

    data.frame(
      iter = i,
      n_atoms = length(z),
      z_min = min(z),
      z_max = max(z),
      J_sum = Jsum,
      J_max = max(J),
      J_max_share = max(Jnorm),
      ess = 1 / sum(Jnorm^2)
    )
  })

  do.call(rbind, out)
}

calc_exceed <- function(z, J, thresholds = c(0.8, 0.9, 0.95)) {
  J <- J / sum(J)
  sapply(thresholds, function(t) sum(J[z >= t]))
}

calc_moments <- function(z, J) {
  J <- J / sum(J)
  m <- sum(z * J)
  v <- sum((z - m)^2 * J)
  c(mean = m, var = v, sd = sqrt(v))
}

## Optional burn-in removal for summaries
burn <- 0
keep <- seq.int(burn + 1, nrow(fit$samples$beta))

## --------------------------
## Beta summary
## --------------------------
beta_samps <- fit$samples$beta[keep, , drop = FALSE]

beta_summary <- data.frame(
  term = colnames(model.matrix(~ biomarker * visit, data = sim_dat)),
  mean = colMeans(beta_samps),
  sd = apply(beta_samps, 2, sd),
  q05 = apply(beta_samps, 2, quantile, 0.05),
  q50 = apply(beta_samps, 2, quantile, 0.50),
  q95 = apply(beta_samps, 2, quantile, 0.95)
)

cat("\n--- Beta posterior summary ---\n")
print(beta_summary)

## If true beta exists
true_beta <- c(0.6, -1.0, -0.8, 0.8)
cat("\n--- Beta comparison ---\n")
print(data.frame(
  term = beta_summary$term,
  truth = true_beta,
  post_mean = beta_summary$mean,
  bias = beta_summary$mean - true_beta
))

## --------------------------
## Rho summary
## --------------------------
rho_samps <- fit$samples$rho[keep]

rho_summary <- data.frame(
  truth = 0.6,
  mean = mean(rho_samps),
  sd = sd(rho_samps),
  q05 = quantile(rho_samps, 0.05),
  q50 = quantile(rho_samps, 0.50),
  q95 = quantile(rho_samps, 0.95)
)

cat("\n--- Rho posterior summary ---\n")
print(rho_summary)

## --------------------------
## CRM summaries
## --------------------------
crm <- fit$samples$crm
z_list <- crm$z.tld[keep]
J_list <- crm$J.tld[keep]

crm_df <- crm_metrics(z_list, J_list)

cat("\n--- CRM summary ---\n")
print(rbind(
  summ_num(crm_df$n_atoms, "n_atoms"),
  summ_num(crm_df$ess, "ess"),
  summ_num(crm_df$J_max_share, "J_max_share"),
  summ_num(crm_df$z_min, "z_min"),
  summ_num(crm_df$z_max, "z_max"),
  summ_num(crm_df$J_sum, "J_sum")
))

cat("\n--- Last 10 CRM iterations ---\n")
print(tail(crm_df, 10))

## --------------------------
## CRM truth comparison
## --------------------------
truth_z <- truth_grid$spt
truth_J <- truth_grid$f0 * truth_grid$dx
truth_J <- truth_J / sum(truth_J)

thresholds <- c(0.8, 0.9, 0.95)

truth_exceed <- calc_exceed(truth_z, truth_J, thresholds)

post_exceed <- t(mapply(
  calc_exceed,
  z = z_list,
  J = J_list,
  MoreArgs = list(thresholds = thresholds)
))

exceed_summary <- data.frame(
  threshold = thresholds,
  truth = as.numeric(truth_exceed),
  post_mean = colMeans(post_exceed),
  post_sd = apply(post_exceed, 2, sd)
)

cat("\n--- Exceedance probabilities ---\n")
print(exceed_summary)

truth_mom <- calc_moments(truth_z, truth_J)

post_mom <- t(mapply(calc_moments, z = z_list, J = J_list))

moment_summary <- data.frame(
  functional = names(truth_mom),
  truth = as.numeric(truth_mom),
  post_mean = colMeans(post_mom),
  post_sd = apply(post_mom, 2, sd)
)

cat("\n--- CRM moment comparison ---\n")
print(moment_summary)

## --------------------------
## Grid-level CRM distance
## --------------------------
grid <- truth_z
breaks <- c(-Inf, (head(grid, -1) + tail(grid, -1)) / 2, Inf)

crm_to_grid <- function(z, J, breaks) {
  J <- J / sum(J)
  bin <- cut(z, breaks = breaks, labels = FALSE, include.lowest = TRUE)

  out <- numeric(length(breaks) - 1)
  rs <- rowsum(J, bin, reorder = FALSE)
  out[as.integer(rownames(rs))] <- rs[, 1]
  out
}

post_grid_mass <- t(mapply(
  crm_to_grid,
  z = z_list,
  J = J_list,
  MoreArgs = list(breaks = breaks)
))

post_grid_mean <- colMeans(post_grid_mass)

crm_distance <- data.frame(
  L1 = sum(abs(post_grid_mean - truth_J)),
  L2 = sqrt(sum((post_grid_mean - truth_J)^2)),
  Linf = max(abs(post_grid_mean - truth_J))
)

cat("\n--- CRM grid distance from truth ---\n")
print(crm_distance)

## --------------------------
## Acceptance rates
## --------------------------
cat("\n--- Acceptance rates ---\n")
print(data.frame(
  beta = fit$p_acc_beta,
  crm = fit$p_acc_crm,
  rho = fit$p_acc_rho
))
```
