---
title: "dirglm_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dirglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dirglm)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will provide a few simulation examples for the `dirglm` function. Each function named `sim_dat_X` generates a data set corresponding to different instances of $\beta$, $f_0$, the link function, or some combination of those three. In what follows, each chunk generates a data set from a `sim_dat_X`, fits `dirglm`, and shows the relevant output.

# Data-Generating Functions
```{r}
# Required Packages ----

require(mvtnorm)
require(gldrm)
require(extraDistr)
require(dplyr)

# Simulation scenario 1 ----
sim_dat_1 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  f0 <- extraDistr::dtpois(spt, lambda = 1, b = 5)     # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                       # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}


# Simulation scenario 2 ----
sim_dat_2 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  temp <- dpois(spt, lambda = 1)
  f0 <- numeric(6)
  f0[1] <- 3 * temp[1] / (3 * temp[1] + sum(temp[-1]))
  f0[-1] <- temp[-1] / (3 * temp[1] + sum(temp[-1]))  # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                        # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}

# Simulation scenario 3 ----
##  Multinomial logistic regression
sim_dat_3 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-1, 1)               # True beta
  spt <- 0:5
  mu <- 5 * plogis(X %*% beta)             # True mu
  f0  <- dbinom(spt, size = 5, prob = 0.5) # True f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}

# Simulation scenario 4 ----
##  Uniform baseline with identity link
sim_dat_4 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), runif(n * (p-1), min=-1, max=1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(5, 2)               # True beta
  mu <- X %*% beta                   # True mu
  spt <- 0:10
  l <- length(spt)
  f0 <- rep(1/l, l)                  # True f0
  theta <- gldrm:::getTheta(
    spt       = spt,
    f0        = f0,
    mu        = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}
```

# Simulation

## Example 1: Truncated Poisson
```{r}
## Example 1
spt <- 0:5
mu0 <- 1
data <- sim_dat_1(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, type="dirglm", dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is truncated Poisson = (0.368, 0.368, 0.184, 0.061, 0.015, 0.003) [rounded]
```

## Example 2: Zero-inflated truncated Poisson
```{r}
## Example 2
spt <- 0:5
mu0 <- 1
data <- sim_dat_2(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, type="dirglm", dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is tilted, truncated Poisson = (0.636, 0.212, 0.106, 0.035, 0.009, 0.002) [rounded]
```

## Example 3: Multiple Logistic Regression
```{r}
## Example 3
spt <- 0:5
mu0 <- 2.5
data <- sim_dat_3(n = 500, p = 3)
### Define link
link <- list(
  linkfun <- function(mu) log(mu / (5 - mu)),
  linkinv <- function(eta) 5 * plogis(eta),
  mu.eta  <- function(eta) 5 * plogis(eta) * (1 - plogis(eta))
)
names(link) <- c("linkfun", "linkinv", "mu.eta")
dir_glm.fit <- dirglm(y ~ X2 + X3, data=data, link=link, type="dirglm",
                      dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-1, 1, 0)
colMeans(f0_samples)          # true f0 is binomial = (0.031, 0.156, 0.313, 0.313, 0.156, 0.031) [rounded]
```

## Example 4: Uniform baseline with Identity link
```{r}
## Example 4
spt <- 0:10
mu0 <- 5
data <- sim_dat_4(n = 500, p = 5)
dir_glm.fit <- dirglm(y ~ X2 + X3 + X4 + X5, data=data, link="identity", type="dirglm",
                        dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(5, 2, 0, 0, 0)
colMeans(f0_samples)          # true f0 is uniform = (0.1, 0.1, 0.1, ..., 0.1)
```

# Some dpglm testing

```{r}
# Simulation III ---------------------------------------------------------------------
f0_true_kde <- readRDS("../DPGLM/DPGLM - only R/Code/sim_truth/f0_true_kde.rds")
spt <- f0_true_kde$spt
f0_kde  <- f0_true_kde$f0 / sum(f0_true_kde$f0)
true_beta <- c(-0.7, 0.2)

true_beta <- c(1, 0)

# A: expit was not written before. Is this right?
expit <- function(x) {
  1 / (1 + exp(-x))
}

sim_I <- function(n) {
  X     <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])
  beta <- true_beta
  mu <- as.numeric(expit(X %*% beta))
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta
  btheta <- b_theta(theta, spt, f0_kde)
  X       <- as.matrix(X)
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })
  # btheta <- theta <- rep(0, n)
  # fY <- t(sapply(1:n, function(i) {
  #   f0_kde
  # }))
  # Y     <- sapply(1:n, function(i) {
  #   sample(spt, 1, prob = fY[i, ])
  # })
  return(data.frame(Y, theta = theta, X))
}


sim_III <- function(n) {
  X     <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])
  beta <- true_beta
  #mu <-  exp(X %*% beta) / (1 + exp(X %*% beta))
  mu <- plogis(X %*% beta)
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta
  
 # theta <- rmvnorm(1, mean = theta, sigma = 2 * diag(n)) %>% as.numeric()
  
  btheta <- b_theta(theta, spt, f0_kde)
  X       <- as.matrix(X)
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })
  return(data.frame(Y, theta = theta, X))
}

truth <- list(beta = true_beta, 
              f0 = matrix(c(spt, f0_kde), 
                          ncol = 2, 
                          byrow = FALSE))

# Simulate Data ---------------------------------------------------------------------
#dat_t <- sim_III(25)
dat_t <- sim_I(250)
y_t <- dat_t[, 1]
#hist(y)
mu0_t <- mean(y_t)
true_theta_t <- dat_t[, 2]
X_t <- dat_t[, -c(1,2)]
datfram_t<- data.frame(X1_t=X_t$X1, X2_t=X_t$X2, y_t=y_t)

model <- dirglm(y_t ~ X2_t, data=datfram_t, link="logit", type="dpglm",
                dpglmControl=dpglm.control(burnin=100, thin=1, save=5000, seed=832))
colMeans(model$samples$beta) # Truth is c(1, 0)
mean(sapply(model$samples$crm$z.tld, mean)) # Truth is 0.5
mean(sapply(model$samples$crm$J.tld, mean)) # Truth is 0.001 (or 0?)
model$p_acc_beta
model$p_acc_f0
```

# Comparison
```{r}
rm(list = ls())

## --------------------------
devtools::load_all()
source("./R/old.R", local = TRUE)

## --------------------------
## Simulation III setup
## --------------------------
f0_true_kde <- readRDS("../DPGLM/DPGLM/sim_truth/f0_kde_unif_0_1.rds")
spt   <- f0_true_kde$spt
f0_kde <- f0_true_kde$f0

dx <- diff(spt)[1]
m0 <- sum(spt * f0_kde) * dx

truth1 <- list(
  beta = c(1, 0),
  f0   = matrix(c(spt, f0_kde), ncol = 2, byrow = FALSE)
)

truth2 <- list(
  beta = c(0.2, 0.7),
  f0   = matrix(c(spt, f0_kde), ncol = 2, byrow = FALSE)
)

## --------------------------
## Data generator
## --------------------------
sim_data <- function(n, true_beta) {
  X <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])

  mu <- as.numeric(stats::plogis(X %*% true_beta))

  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta

  # Compute the matrix of exponentiated outer products
  exp_matrix <- exp(outer(theta, spt, "*"))

  # Compute the dot product row-wise and apply the log
  btheta <- log(exp_matrix %*% f0_kde * dx)

  X <- as.matrix(X)
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })

  return(data.frame(Y, theta = theta, X))
}

## --------------------------
## Run simulation
## --------------------------
iter <- 3000
set.seed(832)
dat1 <- sim_data(250, truth1$beta)
dat_sub <- dat1[1:50, ]

truth1$f0[, 2] <- truth1$f0[, 2] / sum(truth1$f0[, 2])
fit_old <- fit_func(dat_sub, iter, truth1, truth1$beta, 2.5, sqrt(c(1,1)), m0, ebp=2L)

# mb <- truth1$beta
# Sb <- rep(2.5, length(mb))

# crmStart <- list(
#   z.tld = as.numeric(truth1$f0[,1]),
#   J.tld = as.numeric(truth1$f0[,2])
# )
# crmStart$J.tld <- crmStart$J.tld / sum(crmStart$J.tld)

# match fit_func proposal init
# gldrm_fit <- gldrm::gldrm(Y ~ X2, data = data.frame(Y = dat_sub$Y, X2 = dat_sub$X2), link = "logit")
# varbetaStart <- gldrm_fit$varbeta
fit_new <- dirglm(
  Y ~ X2,
  data = data.frame(Y = dat_sub$Y, X2 = dat_sub$X2),
  link = "logit",
  type = "dpglm",
  dpglmControl = dpglm.control(
    burnin = 0, thin = 1, save = iter, seed = 832,
    M = 20#, mu0=m0,
    #mb = mb, Sb = Sb,
    #betaStart = mb,
    #varbetaStart = varbetaStart,
    #thetaStart = dat_sub$theta,
    #crmStart = crmStart
  )
)

out <- list(old = fit_old, new = fit_new)
saveRDS(out, file = "cache/final_sim_dpglm_Jtldonlytilt.rds")
```

```{r}
crm_metrics <- function(zlist, Jlist) {
  stopifnot(length(zlist) == length(Jlist))
  T <- length(zlist)

  out <- data.frame(
    iter = seq_len(T),
    n_atoms = NA_integer_,
    z_min = NA_real_,
    z_max = NA_real_,
    J_sum = NA_real_,
    J_max = NA_real_,
    J_max_share = NA_real_,
    ess = NA_real_
  )

  for (t in seq_len(T)) {
    z <- zlist[[t]]
    J <- Jlist[[t]]
    if (is.null(z) || is.null(J) || length(z) == 0 || length(J) == 0) next

    Jpos <- pmax(J, 0)
    sJ <- sum(Jpos)
    w <- if (sJ > 0) Jpos / sJ else rep(NA_real_, length(Jpos))

    out$n_atoms[t]     <- length(z)
    out$z_min[t]       <- min(z)
    out$z_max[t]       <- max(z)
    out$J_sum[t]       <- sJ
    out$J_max[t]       <- max(Jpos)
    out$J_max_share[t] <- if (sJ > 0) max(w) else NA_real_
    out$ess[t]         <- if (sJ > 0) 1 / sum(w^2) else NA_real_
  }

  out
}

summ_num <- function(x, name = deparse(substitute(x))) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(data.frame(name=name, n=0, min=NA, q05=NA, med=NA, mean=NA, q95=NA, max=NA))
  data.frame(
    name = name,
    n    = length(x),
    min  = min(x),
    q05  = unname(stats::quantile(x, 0.05)),
    med  = stats::median(x),
    mean = mean(x),
    q95  = unname(stats::quantile(x, 0.95)),
    max  = max(x)
  )
}

# out <- readRDS("cache/final_sim_dpglm.rds")
fit_old <- out$old
fit_new <- out$new

# Beta
old_beta_mat <- fit_old$dpglm$beta
old_beta_mean <- colMeans(old_beta_mat, na.rm = TRUE)
old_beta_sd   <- apply(old_beta_mat, 2, sd, na.rm = TRUE)

cat("\n=== OLD model ===\n")
print(data.frame(beta_mean = old_beta_mean, beta_sd = old_beta_sd))
cat("beta_acceptance =", fit_old$dpglm$beta_acceptance, "\n")
cat("crm_acceptance  =", fit_old$dpglm$crm_acceptance, "\n")

new_beta_mat <- fit_new$samples$beta
new_beta_mean <- colMeans(new_beta_mat, na.rm = TRUE)
new_beta_sd   <- apply(new_beta_mat, 2, sd, na.rm = TRUE)

cat("\n=== NEW model ===\n")
print(data.frame(beta_mean = new_beta_mean, beta_sd = new_beta_sd))
cat("beta_acceptance =", fit_new$p_acc_beta, "\n")
cat("crm_acceptance  =", fit_new$p_acc_f0, "\n")

# CRM
old_zlist <- lapply(fit_old$dpglm$crm, `[[`, "z.tld")
old_Jlist <- lapply(fit_old$dpglm$crm, `[[`, "J.tld")
old_crm_df <- crm_metrics(old_zlist, old_Jlist)

new_zlist <- fit_new$samples$crm$z.tld
new_Jlist <- fit_new$samples$crm$J.tld
new_crm_df <- crm_metrics(new_zlist, new_Jlist)

cat("\n--- CRM summary (OLD) ---\n")
print(rbind(
  summ_num(old_crm_df$n_atoms, "n_atoms"),
  summ_num(old_crm_df$ess, "ess"),
  summ_num(old_crm_df$J_max_share, "J_max_share"),
  summ_num(old_crm_df$z_min, "z_min"),
  summ_num(old_crm_df$z_max, "z_max")
))

cat("\n--- CRM summary (NEW) ---\n")
print(rbind(
  summ_num(new_crm_df$n_atoms, "n_atoms"),
  summ_num(new_crm_df$ess, "ess"),
  summ_num(new_crm_df$J_max_share, "J_max_share"),
  summ_num(new_crm_df$z_min, "z_min"),
  summ_num(new_crm_df$z_max, "z_max")
))

# End behavior
cat("\n--- Last 10 CRM iters (OLD) ---\n")
print(tail(old_crm_df, 5))
cat("\n--- Last 10 CRM iters (NEW) ---\n")
print(tail(new_crm_df, 5))
```
