---
title: "dirglm_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dirglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dirglm)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will provide a few simulation examples for the `dirglm` function. Each function named `sim_dat_X` generates a data set corresponding to different instances of $\beta$, $f_0$, the link function, or some combination of those three. In what follows, each chunk generates a data set from a `sim_dat_X`, fits `dirglm`, and shows the relevant output.

# Data-Generating Functions
```{r}
# Required Packages ----

require(mvtnorm)
require(gldrm)
require(extraDistr)
require(dplyr)

# Simulation scenario 1 ----
sim_dat_1 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  f0 <- extraDistr::dtpois(spt, lambda = 1, b = 5)     # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                       # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}


# Simulation scenario 2 ----
sim_dat_2 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = F,
              nrow = n,
              ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-0.7, 0.2)     # true beta
  mu <- exp(X %*% beta)         # true mu, g(mu) = log(mu) = X*beta
  spt <- 0:5                    # support of y
  temp <- dpois(spt, lambda = 1)
  f0 <- numeric(6)
  f0[1] <- 3 * temp[1] / (3 * temp[1] + sum(temp[-1]))
  f0[-1] <- temp[-1] / (3 * temp[1] + sum(temp[-1]))  # true f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0 = f0,
    mu = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta                        # true theta
  prob <- t(sapply(1:n, function(i)
    exp(theta[i] * spt) * f0))
  y <- sapply(1:n, function(i)
    sample(spt, 1, prob = prob[i, ])) # simulated y
  sim.data <- data.frame(X, y)
  return(sim.data)
}

# Simulation scenario 3 ----
##  Multinomial logistic regression
sim_dat_3 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), rnorm(n * (p-1), mean = 0, sd = 1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(-1, 1)               # True beta
  spt <- 0:5
  mu <- 5 * plogis(X %*% beta)             # True mu
  f0  <- dbinom(spt, size = 5, prob = 0.5) # True f0
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}

# Simulation scenario 4 ----
##  Uniform baseline with identity link
sim_dat_4 <- function(n, p, seed=100) {
  set.seed(seed)
  X <- matrix(c(rep(1, n), runif(n * (p-1), min=-1, max=1)),
              byrow = FALSE, nrow = n, ncol = p)
  beta <- numeric(p)
  beta[1:2] <- c(5, 2)               # True beta
  mu <- X %*% beta                   # True mu
  spt <- 0:10
  l <- length(spt)
  f0 <- rep(1/l, l)                  # True f0
  theta <- gldrm:::getTheta(
    spt       = spt,
    f0        = f0,
    mu        = mu,
    sampprobs = NULL,
    ySptIndex = NULL
  )$theta
  prob <- t(sapply(1:n, function(i) exp(theta[i] * spt) * f0))
  # No need to normalize below, prob specified
  y    <- sapply(1:n, function(i) sample(spt, 1, prob = prob[i, ])) # Simulated y
  data.frame(X, y)
}
```

# Simulation

## Example 1: Truncated Poisson
```{r}
## Example 1
spt <- 0:5
mu0 <- 1
data <- sim_dat_1(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, type="dirglm", dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is truncated Poisson = (0.368, 0.368, 0.184, 0.061, 0.015, 0.003) [rounded]
```

## Example 2: Zero-inflated truncated Poisson
```{r}
## Example 2
spt <- 0:5
mu0 <- 1
data <- sim_dat_2(n = 500, p = 2)
dir_glm.fit <- dirglm(y ~ X2, data=data, type="dirglm", dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-0.7, 0.2)
colMeans(f0_samples)          # true f0 is tilted, truncated Poisson = (0.636, 0.212, 0.106, 0.035, 0.009, 0.002) [rounded]
```

## Example 3: Multiple Logistic Regression
```{r}
## Example 3
spt <- 0:5
mu0 <- 2.5
data <- sim_dat_3(n = 500, p = 3)
### Define link
link <- list(
  linkfun <- function(mu) log(mu / (5 - mu)),
  linkinv <- function(eta) 5 * plogis(eta),
  mu.eta  <- function(eta) 5 * plogis(eta) * (1 - plogis(eta))
)
names(link) <- c("linkfun", "linkinv", "mu.eta")
dir_glm.fit <- dirglm(y ~ X2 + X3, data=data, link=link, type="dirglm",
                      dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(-1, 1, 0)
colMeans(f0_samples)          # true f0 is binomial = (0.031, 0.156, 0.313, 0.313, 0.156, 0.031) [rounded]
```

## Example 4: Uniform baseline with Identity link
```{r}
## Example 4
spt <- 0:10
mu0 <- 5
data <- sim_dat_4(n = 500, p = 5)
dir_glm.fit <- dirglm(y ~ X2 + X3 + X4 + X5, data=data, link="identity", type="dirglm",
                        dirglmControl=dirglm.control(mu0=mu0, spt=spt, rho=1))
beta_samples <- dir_glm.fit$samples$beta
f0_samples <- dir_glm.fit$samples$f0
colMeans(beta_samples)        # true beta is c(5, 2, 0, 0, 0)
colMeans(f0_samples)          # true f0 is uniform = (0.1, 0.1, 0.1, ..., 0.1)
```

# Some dpglm testing

```{r}
# Simulation III ---------------------------------------------------------------------
f0_true_kde <- readRDS("../DPGLM/DPGLM - only R/Code/sim_truth/f0_true_kde.rds")
spt <- f0_true_kde$spt
f0_kde  <- f0_true_kde$f0 / sum(f0_true_kde$f0)
true_beta <- c(-0.7, 0.2)

true_beta <- c(1, 0)

# A: expit was not written before. Is this right?
expit <- function(x) {
  1 / (1 + exp(-x))
}

sim_I <- function(n) {
  X     <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])
  beta <- true_beta
  mu <- as.numeric(expit(X %*% beta))
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta
  btheta <- b_theta(theta, spt, f0_kde)
  X       <- X %>% as.matrix()
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })
  # btheta <- theta <- rep(0, n)
  # fY <- t(sapply(1:n, function(i) {
  #   f0_kde
  # }))
  # Y     <- sapply(1:n, function(i) {
  #   sample(spt, 1, prob = fY[i, ])
  # })
  return(data.frame(Y, theta = theta, X))
}


sim_III <- function(n) {
  X     <- cbind(1, runif(n, -sqrt(12) / 4, sqrt(12) / 4))
  X[, -1] <- scale(X[, -1])
  beta <- true_beta
  mu <-  exp(X %*% beta) / (1 + exp(X %*% beta))
  theta <- gldrm:::getTheta(
    spt = spt,
    f0  = f0_kde,
    mu  = mu,
    sampprobs  = NULL,
    ySptIndex  = NULL,
    thetaStart = NULL
  )$theta
  
 # theta <- rmvnorm(1, mean = theta, sigma = 2 * diag(n)) %>% as.numeric()
  
  btheta <- b_theta(theta, spt, f0_kde)
  X       <- X %>% as.matrix()
  fY <- t(sapply(1:n, function(i) {
    exp(theta[i] * spt - btheta[i]) * f0_kde
  }))
  Y     <- sapply(1:n, function(i) {
    sample(spt, 1, prob = fY[i, ])
  })
  return(data.frame(Y, theta = theta, X))
}

truth <- list(beta = true_beta, 
              f0 = matrix(c(spt, f0_kde), 
                          ncol = 2, 
                          byrow = FALSE))

# Simulate Data ---------------------------------------------------------------------
#dat <- sim_III(25)
dat <- sim_I(250)
y <- dat[, 1]
#hist(y)
mu0 <- mean(y)
true_theta <- dat[, 2]
X <- dat[, -c(1,2)]
datfram <- data.frame(X1=X$X1, X2=X$X2, y=y)

model <- dirglm(y ~ X2, data=datfram, link="logit", type="dpglm",
                dpglmControl=dpglm.control(burnin=100, thin=1, save=5000, seed=832, M=200))
colMeans(model$samples$beta) # Truth is c(1, 0)
mean(sapply(model$samples$crm$z.tld, mean)) # Truth is 0.5
mean(sapply(model$samples$crm$J.tld, mean)) # Truth is 0.001 (or 0?)
model$p_acc_beta
model$p_acc_f0
```