---
title: "DPGLM Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dpglm_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The `dirglm` package provides various Bayesian semi-parametric extensions to the generalized linear model. These are
available for use through the functions `dirglm`, `dpglm`, and `cdpglm`.

In this document, we will explore how to use the `dpglm` function and inspect its results. In particular, we will

1. Simulate a simple regression model with continuous, real-valued response
2. Fit `dpglm`
3. Summarize and visualize the results

```{r setup}
library(dirglm)
```

# Simulate Data

We will begin by simulating data from a simple exponential-tilting model.

```{r}
set.seed(832)

n <- 50
x <- rnorm(n)

beta_true <- c(0.2, 0.5)
mu <- plogis(beta_true[1] + beta_true[2] * x)

spt <- seq(1e-4, 1 - 1e-4, length.out = 201)
f0 <- dnorm(spt, mean = 0.55, sd = 0.15)
f0 <- f0 / sum(f0)

theta <- gldrm:::getTheta(
  spt = spt,
  f0 = f0,
  mu = mu,
  sampprobs = NULL,
  ySptIndex = NULL,
  thetaStart = NULL
)$theta

prob_mat <- t(vapply(seq_len(n), function(i) {
  w <- exp(theta[i] * spt) * f0
  w / sum(w)
}, numeric(length(spt))))

y <- vapply(seq_len(n), function(i) {
  sample(spt, size = 1, prob = prob_mat[i, ])
}, numeric(1))

dat <- data.frame(y = y, x = x)
```

# Fit DPGLM

Now, we will fit the `dpglm` model. The number of MCMC draws is kept small to speed up computation, but practitioners may require more iterations for analysis.

```{r}
fit <- dpglm(
  y ~ x,
  data = dat,
  link = "logit",
  dpglmControl = dpglm.control(
    burnin = 10,
    thin = 1,
    save = 50,
    spt = c(0, 1),
    seed = 832
  )
)
```

# Posterior Summary

The `summary()` method reports posterior summaries for the regression coefficients `beta` and reference distribution `crm`.

```{r}
summary(fit)
```

## Plotting

After fitting `dpglm`, one may use `plot_dpglm` to visually inspect some posterior summaries. We will show two examples. The first one inspects `beta`, with some aesthetic customization.

```{r}
plot_dpglm(fit, col = "#E69F00")
```

One can modify the plot aesthetics as shown in the vignette for `dpglm`. Posterior summaries of the reference distribution can be inspected visually, as well.

```{r}
plot_dpglm(fit, what = "crm")
```
